Estimation of flow trajectories in a multi-lines transportation network

### 2023-02-28 latex_files/applied_network_science_article/article_v0.tex
do last corrections

\begin{abstract} % abstract
Characterizing a public transportation network, such as an urban network with multiple lines, requires the origin-destination trip counts during a given period. Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts in vehicles known, it often happens that pedestrian transfers between lines are harder to track, and require costly and invasive devices (e.g., facial recognition system) to be estimated. In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment counts. Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain points in the network, or percentages of transferring passengers between some lines. This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives arguably convincing estimations of the transportation flow.
\end{abstract}

@@

\begin{keyword}
\kwd{multiline bus network}
\kwd{origin-destination flows}
\kwd{boarding and alighting counts}
\kwd{transit flows}
\kwd{maximum entropy estimation}
\kwd{iterative proportional fitting}
\end{keyword}

@@

Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally attempt to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems. In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy. The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for transfer overflow; and an update of marginal distributions. This iterative procedure clearly evokes the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step". The fist step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property). Moreover, this solution offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.

As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network\footnote{\url{https://www.t-l.ch/}} of the city of Lausanne, in Switzerland. Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm. The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a large transportation network and can yield pertinent insights about traveler habits.

Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution. Section \ref{casestudies} contains case studies of the method on toy examples and on the data from the Lausanne public transportation network. Section \ref{conclusion} concludes the article. Data, codes and extensive results can be found in the GitHub repository of the article\footnote{\url{https://github.com/sliunil/tl_study}}.

@@

A \emph{$st$-trip}, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e. achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading to $s$ from $t$.

@@

Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (making this trip do not actually use the line network). The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined following some conditions (see, e.g., section \ref{example_construction}).

@@

\subsection{Statement of the problem and solution method}
\emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}. We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts which can be found in appendix, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (also given in appendix).

Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown. The objective is to estimate the $l\times l$ transportation flow matrix $\mathbf{N}=(n_{st})$. Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}). This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been invoked in transportation modeling researches \cite{wilson1967statistical}  \cite{erlander1990gravity}.

@@

The above iterative solution method is somewhat reminiscent of the EM algorithm. As a matter of fact, the first  step  (maximum entropy) exactly correspond to the ``expectation step" of the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $g_{st}$, $\alpha_s$ and $\beta_t$, do not follow the maximum likelihood rationale of the ``maximisation step". Pseudocode of the algorithm is shown with Algorithm \ref{algo1}.

@@

In view of (\ref{entrer}) and (\ref{sortir}), the postulated margins must satisfy, for each isolated stop $i$
\begin{equation}
\label{ }
\alpha_i=\frac{a_i}{n_{\bullet \bullet}}\qquad\qquad \beta_i=\frac{b_i}{n_{\bullet \bullet}}
\end{equation}
permitting to determine the total flow as $n_{\bullet \bullet}=\frac{a_i}{\alpha_i}$, or  $n_{\bullet \bullet}=\frac{b_i}{\beta_i}$ for any isolated stop $i$, and thus 
the $st$-flow itself as 
\begin{equation}
	\label{flow_from_distrib}
	n_{st} = n_{\bullet \bullet} f_{st}= n_{\bullet \bullet}\phi_s \psi_t g_{st} 
\end{equation}
whose plugging into (\ref{equationGG}) yields the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$. Only the latter is required for subsequent algorithm steps and can be computed with 
\begin{equation}
	z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}
\end{equation}
where $I(.)$ denotes the 0/1 indicator function.
\subsubsection{Initialization of the prior and the margins}
The geometry of the network permits to define the set of permitted $st$-trips across the network denoted by $P$. The initial prior was chosen as the uniform distribution on admissible paths, that is as
\begin{equation*}
g_{st} = \begin{cases}
  \frac{1}{\vert P \vert}    & [s, t] \in $P$, \\
  0    & \text{otherwise}.
\end{cases}
\end{equation*}
The initial margins were chosen to initially match $g_{st}$, namely $\alpha_s=g_{s \bullet}$ and $\beta_t=g_{\bullet t}$ for all stops. However, $g_{st}$ could be chosen more carefully in case of additional data. As a matter of fact, $g_{st}$ represent \emph{prior attractions} between nodes in equation (\ref{Sol}), before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps. An urban planner could chose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.

@@

By construction, the corrected flow $\widetilde{n}_{st}$ found in (\ref{update_flow}) now respect embarkment and disembarkment constraints. We can compute the new transfer flow on edge with
\begin{equation}
\widetilde{z}_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} \widetilde{n}_{st}
\end{equation}
and, with (\ref{entrer}) and (\ref{sortir}), updating margin distributions is straightforward 
\begin{equation}
\label{alpha_beta_update}
\widetilde{\alpha}_s = \frac{a_s - \widetilde{z}_{\bullet s}}{\sum_{s'} (a_{s'} - z_{\bullet {s'}})}   \qquad \qquad \qquad
	\widetilde{\beta}_t = \frac{b_t - \widetilde{z}_{t \bullet}}{\sum_{t'} (b_{t'} - z_{{t'} \bullet})}  \enspace. 
\end{equation}

@@

Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order, that is $F(i)=i+1$. The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips, where $c=\frac{1}{(l-1)(l-2)}$.   The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known
\begin{equation}
\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet\enspace. 
\end{equation}
Following  (\ref{Sol}) maximum entropy flows are of the form
\begin{equation}
\label{nosignle}
n_{st}= n_{\bullet\bullet}\,  c\, I(s<t)\, \phi_s\,  \psi_t 
\end{equation}
where (setting $\Psi_s:=\sum_{t>s}\psi_t$ and $\Phi_t:=\sum_{s<t}c\phi_s$) the constraints (\ref{Sol_LagMult}) equivalently read
\begin{equation}
\label{dis embarking constraints}
\phi_s=\frac{\alpha_s}{c\, \sum_{t>s}\psi_t}=\frac{a_s}{n_{\bullet\bullet}\,  c\, \Psi_s}
\qquad\qquad\qquad
\psi_t=\frac{\beta_t}{c\, \sum_{s<t}\phi_s}=\frac{b_t}{n_{\bullet\bullet}\, c\, \Phi_t}
\end{equation}
to be solved by iterative fitting.

@@

This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the bus at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$. By (\ref{nosignle}), 
\begin{displaymath}
m_{st}=\sum_{u\ge t}n_{su}=n_{\bullet\bullet}\, c\, \phi_s  \sum_{u\ge t} I(s<u)\, \psi_u =n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t) 
\end{displaymath}
The empirical estimate of $\rho_{st}$ is given by the proportion, among the travelers embarking at $s$ and 
 still present at $t>s$,  of travelers disembarking at $t$, that is 
\begin{displaymath}
\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{n_{\bullet\bullet}\, c\,  \phi_s\,   \psi_t}{n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t)}=\frac{\psi_t}{\psi_t+\Psi_t}\le 1 
\end{displaymath}
which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{\psi_t}{\psi_t+\Psi_t}$ at $t$ is {\em independent} of the embarkment stop $s$. Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.

@@

Case studies are divided in two sections. In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is randomly drawn. These examples enable some kind of validation of the algorithm, as the "real" transportation flow is known and can be compared to the solution $\mathbf{N} = (n_{st})$ given by our method. This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but the real transportation flow is unknown. This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.

@@

In all case studies, we obtain a estimation of the transportation flow with the algorithm, noted $\mathbf{N} = (n_{st})$, starting from the real embarkment flow $\mathbf{a}_\text{ref}$ and disembarkment flow $\mathbf{b}_\text{ref}$. In toy examples, we also have access to the real transportation flow $\mathbf{N}_\text{ref}$. 
There are two types of dissimilarity measures between the data and the solution proposed by the algorithm: (1) if we have access to $\mathbf{N}_\text{ref}$, how much $\mathbf{N}$ differs from it, and (2) how well constraints defined by $\mathbf{a}_\text{ref}$ and $\mathbf{b}_\text{ref}$ are respected. The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed as
\begin{equation}
	\text{MTE}(\mathbf{N}) = \sum_{st} \frac{n^\text{ref}_{st}}{n^\text{ref}_{\bullet \bullet}} \frac{\lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{st}} = \frac{\sum_{st} \lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{\bullet \bullet}} 
	\label{MTE}
\end{equation} 
and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, defined as
\begin{align}
	\text{MME}(\mathbf{N}) &= \frac{1}{2} \sum_{i} \frac{a^\text{ref}_i}{a^\text{ref}_\bullet} \frac{\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert}{a^\text{ref}_i} + \frac{1}{2} \sum_{i} \frac{b^\text{ref}_i}{b^\text{ref}_\bullet} \frac{\lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert}{a^\text{ref}_i} \notag \\
	&= \frac{\sum_i (\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert + \lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert)}{2n^\text{ref}_{\bullet \bullet}}
	\label{MME}
\end{align} 
where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean of a percentage of error.

@@

All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic. We fix a number of \emph{round trips} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines. Every line has a starting and ending node, which are isolated nodes, and possesses $p-1$ intermediary nodes which allow transfers to the other round trips, yielding a total of $n = 2p(p + 1)$ nodes in the network. Examples of these toy networks can be found in Figure \ref{toy_example_plots}.

@@

First, we exhibit some algorithm iterations on a toy example with $p=2$, where $50$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips. Some iterations of the algorithm with $\theta=0.001$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}. 
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ round trips, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$. MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
	\label{iteration_plots}
\end{figure}
On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.256. This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximizes the entropy.

@@

The randomness of the $st$-trips distribution is controlled by the number of drawn passengers and we can see that the algorithm performs better if this number increases, as seen in Figure \ref{MTE_vs_passengers}. This behavior can be expected as the method is constructed to find the solution maximizing the entropy while respecting embarkment and disembarkment constraints. In fact, all MTE should converge to $0$ eventually, however this rate of convergence seems to be low.
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
	\caption{Mean Transportation Error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of round trips $p$ varying from 2 to 8. The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
	\label{MTE_vs_passengers}
\end{figure}
As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decrease as the network size increases, reaching value close to $0$ for $8$ round trips. However, this is not the only factor which helps to calibrate this parameter. As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
	\caption{Mean Transportation Error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of round trips $p$ varying from 2 to 8. The dot on each curve denotes the minimum. The number of drawn passengers is different for every network size and is set to $5 \cdot \vert P \vert$, in order to keep the randomness level constant. $10$ different draws are performed for each data point.}
	\label{MTE_vs_theta}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ by ``superstop" names.}
	\label{transfers}
\end{figure}

>>>>

\begin{abstract} % abstract
Characterizing a public transportation network, such as an urban network with multiple lines, requires the origin-destination trip counts during a given period. Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts in vehicles known, it often happens that pedestrian transfers between lines are harder to track, and require costly and invasive devices (e.g., facial recognition system) to be estimated. In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment counts. Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain nodes in the network, or percentages of transferring passengers between some lines. This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives arguably convincing estimations of the transportation flow.
\end{abstract}

@@

\begin{keyword}
\kwd{multiline bus network}
\kwd{origin-destination flows}
\kwd{boarding and alighting counts}
\kwd{maximum entropy estimation}
\kwd{iterative proportional fitting}
\end{keyword}

@@

Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally attempt to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems. In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy. The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for transfer overflow; and an update of marginal distributions. This iterative procedure clearly evokes the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step". The first step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property). This method also offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.

As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network\footnote{\url{https://www.t-l.ch/}} of the city of Lausanne, in Switzerland. Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm. The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a real transportation network and can yield pertinent insights about traveler habits.

Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution. Section \ref{casestudies} contains case studies with toy examples and the Lausanne public transportation network. Section \ref{conclusion} concludes the article. Data, codes and extensive results can be found in the GitHub repository of the article\footnote{\url{https://github.com/sliunil/tl_study}}.

@@

A \emph{$st$-trip}, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e. achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading from $s$ to $t$.

@@

Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (travelers making this trip do not actually do not use the transportation network). The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined following some conditions (see, e.g., section \ref{example_construction}).

@@

\subsection{Statement of the problem and solution method}
\emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}. We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (given in the appendix).

Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown. The objective is to estimate the $l\times l$ transportation flow matrix $\mathbf{N}=(n_{st})$. Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}). This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been invoked in transportation modeling research \cite{wilson1967statistical}  \cite{erlander1990gravity}.

@@

The above iterative solution method is somewhat reminiscent of the EM algorithm. As a matter of fact, the first  step  (maximum entropy) exactly correspond to the ``expectation step" of the EM algorithm (see e.g. \cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $g_{st}$, $\alpha_s$ and $\beta_t$, do not follow the maximum likelihood rationale of the ``maximisation step". Pseudocode of the algorithm is shown in Algorithm \ref{algo1}.

@@

In view of (\ref{entrer}) and (\ref{sortir}), the postulated margins must satisfy, for each isolated stop $i$
\begin{equation}
\label{ }
\alpha_i=\frac{a_i}{n_{\bullet \bullet}}\qquad\qquad \beta_i=\frac{b_i}{n_{\bullet \bullet}}
\end{equation}
permitting to determine the total flow as $n_{\bullet \bullet}=\frac{a_i}{\alpha_i}$, or  $n_{\bullet \bullet}=\frac{b_i}{\beta_i}$ for any isolated stop $i$, and thus 
the $st$-flow itself as 
\begin{equation}
	\label{flow_from_distrib}
	n_{st} = n_{\bullet \bullet} f_{st}= n_{\bullet \bullet}\phi_s \psi_t g_{st} 
\end{equation}
whose plugging into (\ref{equationGG}) yields the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$. Only the latter is required for subsequent algorithm steps and can be computed with 
\begin{equation}
	z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}
\end{equation}
where $I(.)$ denotes the 0/1 indicator function.
\subsubsection{Initialization of the prior and the margins}
The geometry of the network permits to define the set of permitted $st$-trips across the network denoted by $P$. The initial prior was chosen as the uniform distribution on admissible paths, that is as
\begin{equation*}
g_{st} = \begin{cases}
  \frac{1}{\vert P \vert}    & [s, t] \in $P$, \\
  0    & \text{otherwise}.
\end{cases}
\end{equation*}
The initial margins were chosen to initially match $g_{st}$, namely $\alpha_s=g_{s \bullet}$ and $\beta_t=g_{\bullet t}$ for all stops. However, $g_{st}$ could be chosen more carefully in case of additional data. As a matter of fact, $g_{st}$ represent \emph{prior attractions} between nodes in equation (\ref{Sol}), before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps. An urban planner could choose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.

@@

By construction, the corrected flow $\widetilde{n}_{st}$ found in (\ref{update_flow}) now respects embarkment and disembarkment constraints. We can compute the new transfer flow on edges with
\begin{equation}
\widetilde{z}_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} \widetilde{n}_{st}
\end{equation}
and, with (\ref{entrer}) and (\ref{sortir}), updating margin distributions is straightforward 
\begin{equation}
\label{alpha_beta_update}
\widetilde{\alpha}_s = \frac{a_s - \widetilde{z}_{\bullet s}}{\sum_{s'} (a_{s'} - z_{\bullet {s'}})}   \qquad \qquad \qquad
	\widetilde{\beta}_t = \frac{b_t - \widetilde{z}_{t \bullet}}{\sum_{t'} (b_{t'} - z_{{t'} \bullet})}  \enspace. 
\end{equation}

@@

Let  $i=1,\ldots, l$ enumerate the stops in increasing order, that is $F(i)=i+1$. The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips, where $c=\frac{1}{(l-1)(l-2)}$.   The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known
\begin{equation}
\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet\enspace. 
\end{equation}
Following  (\ref{Sol}) maximum entropy flows are of the form
\begin{equation}
\label{nosignle}
n_{st}= n_{\bullet\bullet}\,  c\, I(s<t)\, \phi_s\,  \psi_t 
\end{equation}
where (setting $\Psi_s:=\sum_{t>s}\psi_t$ and $\Phi_t:=\sum_{s<t}c\phi_s$) the constraints (\ref{Sol_LagMult}) equivalently read
\begin{equation}
\label{dis embarking constraints}
\phi_s=\frac{\alpha_s}{c\, \sum_{t>s}\psi_t}=\frac{a_s}{n_{\bullet\bullet}\,  c\, \Psi_s}
\qquad\qquad\qquad
\psi_t=\frac{\beta_t}{c\, \sum_{s<t}\phi_s}=\frac{b_t}{n_{\bullet\bullet}\, c\, \Phi_t}
\end{equation}
to be solved by iterative fitting.

@@

This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the line at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$. By (\ref{nosignle}), 
\begin{displaymath}
m_{st}=\sum_{u\ge t}n_{su}=n_{\bullet\bullet}\, c\, \phi_s  \sum_{u\ge t} I(s<u)\, \psi_u =n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t) 
\end{displaymath}
The empirical estimate of $\rho_{st}$ is given by the proportion, among the travelers embarking at $s$ and 
 still present at $t>s$,  of travelers disembarking at $t$, that is 
\begin{displaymath}
\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{n_{\bullet\bullet}\, c\,  \phi_s\,   \psi_t}{n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t)}=\frac{\psi_t}{\psi_t+\Psi_t}\le 1 
\end{displaymath}
which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{\psi_t}{\psi_t+\Psi_t}$ at $t$ is {\em independent} of the embarkment stop $s$. Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the line at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.

@@

Case studies are divided in two sections. In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is randomly drawn. These examples enable some kind of validation of the algorithm, as the "real" transportation flow is known and can be compared to the solution $\mathbf{N} = (n_{st})$ given by our method. This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but the real transportation flow is unknown. This second case study shows that the algorithm is applicable on large, real datasets and can give insights about passengers probable routes in the network.

@@

In all case studies, we obtain a estimation of the transportation flow with the algorithm, noted $\mathbf{N} = (n_{st})$, starting from the real embarkment flow $\mathbf{a}_\text{ref}$ and disembarkment flow $\mathbf{b}_\text{ref}$. In toy examples, we also have access to the real transportation flow $\mathbf{N}_\text{ref}$. 
There are two types of dissimilarity measures between the data and the solution proposed by the algorithm: (1) if we have access to $\mathbf{N}_\text{ref}$, how much $\mathbf{N}$ differs from it, and (2) how well constraints defined by $\mathbf{a}_\text{ref}$ and $\mathbf{b}_\text{ref}$ are respected. The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed as
\begin{equation}
	\text{MTE}(\mathbf{N}) = \sum_{st} \frac{n^\text{ref}_{st}}{n^\text{ref}_{\bullet \bullet}} \frac{\lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{st}} = \frac{\sum_{st} \lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{\bullet \bullet}} 
	\label{MTE}
\end{equation} 
and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, defined as
\begin{align}
	\text{MME}(\mathbf{N}) &= \frac{1}{2} \sum_{i} \frac{a^\text{ref}_i}{a^\text{ref}_\bullet} \frac{\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert}{a^\text{ref}_i} + \frac{1}{2} \sum_{i} \frac{b^\text{ref}_i}{b^\text{ref}_\bullet} \frac{\lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert}{a^\text{ref}_i} \notag \\
	&= \frac{\sum_i (\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert + \lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert)}{2n^\text{ref}_{\bullet \bullet}}
	\label{MME}
\end{align} 
where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean of error percentages.

@@

All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic. We fix a number of \emph{round trips} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines. Every line has a starting and an ending node, which are isolated nodes, and possesses $p-1$ intermediary nodes which allow transfers to the other round trips, yielding a total of $n = 2p(p + 1)$ nodes in the network. Examples of these toy networks can be found in Figure \ref{toy_example_plots}.

@@

First, we exhibit some algorithm iterations on a toy example with $p=2$, where $50$ passengers were drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips. Some iterations of the algorithm with $\theta=0.001$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}. 
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ round trips, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$. MTE and MME errors are computed, and embarkment and disembarkment counts are represented respectively by the red and blue colors on nodes.}
	\label{iteration_plots}
\end{figure}
On this small example, we can see that the algorithm quickly find an estimation giving a small MME error, but still give a MTE of 0.256. This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximizes the entropy.

@@

The randomness of the $st$-trips distribution is controlled by the number of drawn passengers and we can see that the algorithm performs better if this number increases, as seen in Figure \ref{MTE_vs_passengers}. This behavior can be expected as the method is constructed to find the solution maximizing the entropy while respecting embarkment and disembarkment constraints. In fact, all MTE should converge to $0$ eventually, however this rate of convergence seems to be low.
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
	\caption{Mean transportation error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of round trips $p$ varying from 2 to 8. The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
	\label{MTE_vs_passengers}
\end{figure}
As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decrease as the network size increases, reaching value close to $0$ for $8$ round trips. However, this is not the only factor which helps to calibrate this parameter. As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
\begin{figure}[h]
	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
	\caption{Mean transportation error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of round trips $p$ varying from 2 to 8. The dot on each curve denotes the minimum. The number of drawn passengers is different for every network size and is set to $5 \cdot \vert P \vert$, in order to keep an (almost) constant variance. $10$ different draws are performed for each data point.}
	\label{MTE_vs_theta}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
	\caption{Localization and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ in ``superstops".}
	\label{transfers}
\end{figure}
